---
title: "Untitled"
author: "Ivan García"
date: "2025-05-05"
output: html_document
---

DAGS: https://www.statistical-rethinking.robitalec.ca/week-7.html

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Paso 1: Preparar el entorno y los datos
Primero cargamos los paquetes necesarios y el dataset.

```{r}
library(rethinking)
data(Trolley)
d <- Trolley
```

## Histograma

```{r}
simplehist( d$response , xlim=c(1,7) , xlab="response" )
```

## Cumulative probabilities

```{r}
# discrete proportion of each response value
pr_k <- table( d$response ) / nrow(d)
# cumsum converts to cumulative proportions
cum_pr_k <- cumsum( pr_k )
# plot
plot( 1:7 , cum_pr_k , type="b" , xlab="response" ,
ylab="cumulative proportion" , ylim=c(0,1) )
```

## Intercept parameters 

```{r}
logit <- function(x) log(x/(1-x)) # convenience function
( lco <- logit( cum_pr_k ) )
```

## Basic model (no predictor variables) fitting

```{r}
m11.1 <- map(
alist(
response ~ dordlogit( phi , c(a1,a2,a3,a4,a5,a6) ),
phi <- 0,
c(a1,a2,a3,a4,a5,a6) ~ dnorm(0,10)
) ,
data=d ,
start=list(a1=-2,a2=-1,a3=0,a4=1,a5=2,a6=2.5) )
```


```{r}
precis(m11.1)
```

## Get cumulative probabilities back

```{r}
logistic(coef(m11.1))
```

Same as the values in cum_pr_k that we computed earlier. But
now we also have a posterior distribution around these values, and we’re ready to add pre-
dictor variables in the next section.

```{r}
# note that data with name 'case' not allowed in Stan
# so will pass pruned data list
m11.1stan <- ulam(
  alist(
    response ~ dordlogit(0, cutpoints),  # phi = 0 (modelo nulo)
    cutpoints ~ dnorm(0, 10)
  ),
  data = list(response = d$response),
  start = list(cutpoints = c(-2, -1, 0, 1, 2, 2.5)),
  chains = 2, 
  cores = 2
)

# Para ver los resultados
precis(m11.1stan, depth = 2)

```
## Now we include predictor variables
```{r}
m11.2_stan <- ulam(
  alist(
    response ~ dordlogit(phi, cutpoints),
    phi <- bA * action + bI * intention + bC * contact,
    c(bA, bI, bC) ~ dnorm(0, 10),
    cutpoints ~ dnorm(0, 10)
  ),
  data = d,  # Mismo dataframe con las variables necesarias
  start = list(cutpoints = c(-1.9, -1.2, -0.7, 0.2, 0.9, 1.8)),
  chains = 2,  # Cadenas MCMC
  cores = 2,   # Paralelización
  iter = 1000  # Iteraciones por cadena
)

# Resumen del modelo
precis(m11.2_stan, depth = 2)
```
## m11.3 Adds two interaction terms and two interaction parameters, bAI and bCI.

```{r}
m11.3_stan <- ulam(
  alist(
    response ~ dordlogit(phi, cutpoints),
    phi <- bA * action + bI * intention + bC * contact +
           bAI * action * intention + bCI * contact * intention,
    c(bA, bI, bC, bAI, bCI) ~ dnorm(0, 10),
    cutpoints ~ dnorm(0, 10)  # Puntos de corte como vector
  ),
  data = d,
  start = list(cutpoints = c(-1.9, -1.2, -0.7, 0.2, 0.9, 1.8)),
  chains = 2,  # Número de cadenas MCMC
  cores = 2,   # Paralelización
  iter = 1000  # Iteraciones por cadena
)

# Resumen del modelo
precis(m11.3_stan, depth = 2)
```

Now let’s compare these three models

```{r}

# Instalar si es necesario (solo una vez)
# install.packages("rethinking")

library(rethinking)

# Usar precis() para cada modelo y combinar manualmente
precis_m11.1 <- precis(m11.1stan, depth = 2)
precis_m11.2 <- precis(m11.2_stan, depth = 2)
precis_m11.3 <- precis(m11.3_stan, depth = 2)

# Combinar resultados en una lista
combined_results <- list(
  m11.1 = precis_m11.1,
  m11.2 = precis_m11.2,
  m11.3 = precis_m11.3
)

# Imprimir comparación
print(combined_results)

```
WAIC
```{r}
# Opción 1: Usar LOO-CV (recomendado)
loo_compare(loo(m11.1stan), loo(m11.2_stan), loo(m11.3_stan))

# Opción 2: Usar WAIC (alternativa clásica)
compare(m11.1, m11.2, m11.3, func = WAIC)

# Opción 3: Comparación visual con bayesplot
library(bayesplot)
posterior <- cbind(
  as.matrix(m11.1)[,c("bA","bI","bC")],
  as.matrix(m11.2)[,c("bA","bI","bC")],
  as.matrix(m11.3)[,c("bA","bI","bC")]
)
mcmc_intervals(posterior)
```

